---
title: "DataAnalysis"
output: html_document
---
#**BIOL 812 Assignment 2: Part II**

##Load Packages and Data

  The first step to our analysis is loading the required packages (dplyr, ggplot2, knitr) and the data. There will be some warnings when loading the R packages, as we are using an older version of R. We call the R object storing our data "MData". The first column of the .csv gives the row numbers (1 through 10) and as such is not needed. Thus, we take all columns except the first from our .csv that contains the data, and load them to the object Mdata.

```{r Load Packages and Data}

library(dplyr)
library(ggplot2)
library(knitr)
MData<-read.csv("Data/measurements.csv")[,-1]
```

###About the data...

  There are five columns in MData. The first is Organism, which tells us the species for a given row. Next, we have Limb.Width and unitsW, which provide the limb width and the units that the width was measured in, for each species. Finally, we have Limb.Length and unitsL, which provide the limb length and the units that the length was measured in, for each species.

##Data Analysis

  Our data analysis is consists of six sections:

  * Changing column names to 3 letter abbreviations
  * Calculating limb volume from limb width and limb length
  * Grouping the species into 2 groups
  * Calculating summary statistics (mean, standard deviation)
  * Plotting limb width (y) against limb length (x)
    * Plotting limb width (y) against limb length (x), for species with above-average volume
  * Creating a table showing species and volume for the species with below-average volume

###Rename Columns

  We use the rename function from dplyr to change the column names to 3 letter abbreviations (Org, Lwd, Uwd, Lln and Uln for organism, limb width, units of limb width, limb length and units of limb length, respectively).

```{r Rename Columns}

MData<-rename(MData,Org=Organism,Lwd=Limb.Width,Uwd=unitsW,Lln=Limb.Length,Uln=unitsL)
```

###Calculate Volume

  The first step to calculating limb volume for each species is to ensure that all of our measurements are in the same units. Here, we will use centimeters as our unit of measure. All measurements in millimeters must be divided by 10, to convert to centimeters. This conversion is completed using the functions widthtocm and lengthtocm for limb width and limb length, respectively. We add a column to our dataframe with measurements in centimeters for both limb width and limb length, using the mutate function from dplyr. Remark that the units for volume are cm^3.

  Once all of the measurements are in centimeters, we obtain our limb volume using the formula $V=\pi*r^2*h$, where $r$ is radius and $h$ is height. Remark that this is the formula for the volume of a cylinder, which we have selected to calculate limb volumes as we assume that the shape of limbs are approximately cylindrical. We take limb width to be the diameter (two times the radius) of the limb, and limb length to be the height in our volume formula. We add this limb volume as a column titled "Volume" in our dataframe using the mutate function from dplyr. 

```{r Calculate Volume}

widthtocm<-function(i){
if(MData$Uwd[i]=="mm"){
  return(MData$Lwd[i]/10)}
  else{return(MData$Lwd[i])}}
MData<-mutate(MData,Width.All.cm=sapply(1:nrow(MData),widthtocm))

lengthtocm<-function(i){
if(MData$Uln[i]=="mm"){
  return(MData$Lln[i]/10)}
  else{return(MData$Lln[i])}}
MData<-mutate(MData,Length.All.cm=sapply(1:nrow(MData),lengthtocm))

MData<-mutate(MData,Volume=pi*((Width.All.cm/2)^2)*Length.All.cm)
```

###Assign Groups

  We assign our species to one of two groups: salt water and fresh water. This assignment is random, with a probability of 0.5 for each of salt and fresh water. The column of groups is added to our dataframe using the mutate function from dplyr. We set a seed to ensure reproducibility of the groups assigned in our data analysis.
  
```{r Assign Groups}

set.seed(777)
MData<-mutate(MData,Group=sample(c("Salt.Water","Fresh.Water"),size=10,replace=TRUE,prob=c(0.5,0.5)))
```

###Calculate Summary Statistics by Group

  We calculate the mean and standard deviation of limb width, limb length and limb volume for each of our groups (salt water and fresh water). The means and standard deviations are calculated for each group using the group_by function to separate our species by the group (salt or fresh water) that they belong to. These values are placed into a table, using the summarize function in dplyr to ensure one row per group (one row for means and standard deviations of fresh water species, and one row for those of salt water species). The kable function from the knitr package is used to make our final table more presentable in the report.

```{r Calculate Summary Statistics by Group}
kable(summarize(group_by(MData,Group),Mean.Width=mean(Width.All.cm),SD.Width=sd(Width.All.cm),Mean.Length=mean(Length.All.cm),SD.Length=sd(Length.All.cm),Mean.Volume=mean(Volume),SD.Volume=sd(Volume)), caption="Table 1: Table of summary statistics.")
```

###Plot Limb Width (y) Against Length (x)

  We use ggplot to create a scatterplot of limb width (y) against limb length (x) for our ten species. Thus, there is one point for each species. The points are coloured by the group that the species belongs to (fresh water or salt water).
  
```{r Plot Limb Width (y) Against Length (x)}
ggplot(MData,aes(Length.All.cm,Width.All.cm,colour=Group))+
  geom_point(size=I(4),alpha=0.7)+
  xlab("Limb length (cm)")+
  ylab("Limb width (cm)")+
  labs(colour="Group")+
  scale_colour_discrete(name="Group",labels=c("Fresh water","Salt water"))+
  theme_light()
```

####Plot Limb Width (y) Against Length (x) for Species with Above-Average Volume

  To plot limb width (y) against limb length (x) for only the species with above-average volume, we first generate a new data frame that contains only the species with above-average volume. We call the object storing this dataframe MDataLg. We take average volume to be the mean volume. The group_by function from dplyr is used to group the species by their group (fresh water or salt water), and the filter function from dplyr is used to keep only the rows with species that have volume greater than the average.
  
  Again, we use ggplot to create a scatterplot of limb width (y) against limb length (x) for the species with above-average volume. The points are coloured by the group that the species belongs to (fresh water or salt water).

```{r}
#Mean volume is 696.13 for freshwater and 647.83 for salt water
MDataLg<-filter(group_by(MData,Group),Volume>mean(Volume))

ggplot(MDataLg,aes(Length.All.cm,Width.All.cm,colour=Group))+
  geom_point(size=I(4),alpha=0.7)+
  xlab("Limb length (cm)")+
  ylab("Limb width (cm)")+
  labs(colour="Group")+
  scale_colour_manual(values=c("darkgreen","goldenrod1"),name="Group",labels=c("Fresh water","Salt water"))+
  theme_light()
```

###Table of Species and Volume

  For the last part of our analysis, we generate a table of species and volume for the species with below average volume. We repeat the process (using the group_by and filter functions) that was used to generate the MDataLg dataframe above to generate the MdataSm dataframe, which contains only species with a volume that is lower than the mean volume. We then ungroup our dataframe using the ungroup function in the dplyr package, and select only the column containing species name and the column containing volume to appear in our table. Note that we rename our "Org" column to "Species", so that the column of the table containing species is labelled more appropriately. The kable function from the knitr package is used to make our final table more presentable in the report.

```{r Table}
MdataSm<-filter(group_by(MData,Group),Volume<mean(Volume))
kable(select(rename(ungroup(MdataSm),Species=Org),Species,Volume),caption="Table 2: Table of species and their volumes for species with below-average volume.")
```
